\Large The state learner -- a super learner for right-censored data

### 2024-12-20 manuscript/manuscript.tex
adapted Thomas changes

There are many existing super learners for right-censored data. We
briefly review them here. Machine learning based on right-censored
data commonly uses the partial log-likelihood as a loss function
\citep[e.g.,][]{li2016regularized,yao2017deep,lee2018deephit,katzman2018deepsurv,gensheimer2019scalable,lee2021boosted,kvamme2021continuous}.
This loss function is also suggested for super learning with
right-censored data by \cite{polley2011-sl-cens}, where data is
assumed to be observed in discrete time. However, the partial
log-likelihood loss does not work well with data splitting
(cross-validation) in continuous time. The reason is that the partial
log-likelihood loss assigns an infinite value for any learner that
predicts piece-wise constant cumulative hazard functions when there
are event times in the test set which do not occur in the learning
set. This problem occurs with prominent survival learners including
the Kaplan-Meier estimator, the random survival forest, and the
semi-parametric Cox regression model and these learners cannot be
included in the library of the super learner proposed by
\cite{polley2011-sl-cens}. When a proportional hazards model is
assumed, the baseline hazard function can be profiled out of the
likelihood \citep{cox1972regression}. The cross-validated partial
log-likelihood loss \citep{verweij1993cross} has therefore been
suggested as a loss function for super learning
\citep{golmakani2020super}. This choice of loss function restricts the
library of learners to include only Cox proportional hazards models,
and hence excludes many learners such as, e.g., random survival
models, additive hazards models and accelerated failure time models.
Other proposals for super learning with right-censored data use an
inverse probability of censoring weighted (IPCW) loss function
\citep{graf1999assessment,van2003unicv,molinaro2004tree,keles2004asymptotically,hothorn2006survival,gerds2006consistent,gonzalez2021stacked},
censoring unbiased transformations
\citep{fan1996local,steingrimsson2019censoring}, or pseudo-values
\citep{andersen2003generalised,mogensen2013random,sachs2019ensemble}.
All these methods rely on an estimator of the censoring distribution,
and their drawback is that this estimator has to be pre-specified. An
approach which avoids a pre-specified censoring model was proposed
independently by \cite{han2021inverse} and
\cite{westling2021inference}. In both articles, the authors suggest to
iterate between learning of the outcome model and learning of the
censoring model using IPCW loss functions. However, no general
theoretical guarantees seem to exist for this procedure and the
procedure is not guaranteed to converge to the true data-generating
mechanism \citep{munch2024thesis}.

>>>>

There are many existing super learners for right-censored data. We briefly
review them here. Machine learning based on right-censored data commonly uses
the partial log-likelihood as a loss function
\citep[e.g.,][]{li2016regularized,yao2017deep,lee2018deephit,katzman2018deepsurv,gensheimer2019scalable,lee2021boosted,kvamme2021continuous}.
This loss function is also suggested for super learning with right-censored data
by \cite{polley2011-sl-cens}, where data is assumed to be observed in discrete
time. However, the partial log-likelihood loss does not work well with data
splitting (cross-validation) in continuous time. The reason is that the partial
log-likelihood loss assigns an infinite value for any learner that predicts
piece-wise constant cumulative hazard functions when there are event times in
the test set which do not occur in the learning set. This problem occurs with
prominent survival learners including the Kaplan-Meier estimator, the random
survival forest, and the semi-parametric Cox regression model and these learners
cannot be included in the library of the super learner proposed by
\cite{polley2011-sl-cens}. When a proportional hazards model is assumed, the
baseline hazard function can be profiled out of the likelihood
\citep{cox1972regression}. The cross-validated partial log-likelihood loss
\citep{verweij1993cross} has therefore been suggested as a loss function for
super learning by \cite{golmakani2020super}. This choice of loss function
restricts the library of learners to include only Cox proportional hazards
models, and hence excludes many learners such as, e.g., random survival models,
additive hazards models, and accelerated failure time models. Other proposals
for super learning with right-censored data use an inverse probability of
censoring weighted (IPCW) loss function
\citep{graf1999assessment,van2003unicv,molinaro2004tree,keles2004asymptotically,hothorn2006survival,gerds2006consistent,gonzalez2021stacked},
censoring unbiased transformations
\citep{fan1996local,steingrimsson2019censoring}, or pseudo-values
\citep{andersen2003generalised,mogensen2013random,sachs2019ensemble}. All these
methods rely on an estimator of the censoring distribution, and their drawback
is that this estimator has to be pre-specified. An approach which avoids a
pre-specified censoring model was proposed independently by
\cite{han2021inverse} and \cite{westling2021inference}. In both articles, the
authors suggest to iterate between learning of the outcome model and learning of
the censoring model using IPCW loss functions. However, it seems that this
procedure is in general not guaranteed to converge to the true data-generating
mechanism \citep{munch2024thesis}.
